The Nuclear Norm and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-08-19 conf_article/neon.tex
Updates included

%------------------------------------------------------------------------------------
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{op}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{op}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

\usepackage{amsmath}
\DeclareMathOperator{\tr}{tr}

@@

\subsection{Linear Minimization Oracle}
Let us look at the problem from the perspective of linear minimization oracle (lmo) and unconstrained stochastic conditional gradient descent (uSCG) (\cite{pethick2025training}). lmo is defined as:
\begin{equation}\label{eq:lmo}
    \mathrm{lmo}(\mS) \in \arg\min_{\mX \in \cS} \<\mS, \mX>,
\end{equation}
where $\cS$ is some set. We are interested in the case when $\cS$ is a ball in our $\norm{\cdot}$ norm:
\begin{equation}\label{eq:ball}
    \cS := \cB_\eta :=\{ \mX \mid \norm{\mX} \leq \eta \}.
\end{equation}

@@

It can be easily shown that the formula is equvalent to
\begin{equation}\label{eq:trust_update}
    \mX^{k+1} = \mX^{k} - \gamma_k\eta \arg\max_{\mX \in \cB_1} \<\mS, \mX> = \mX^{k} - \gamma_k\eta \{\Delta \in \cB_1 \mid \<\mM^k, \Delta> = \norm{\mM^k}^\dagger\}.
\end{equation}

Let us set $\gamma_k = 1$. This transforms algorithm defined by \cref{eq:trust_update} into Algorithm 1 from \cite{kovalev2025understanding}. Therefore, we can view \cref{eq:trust_update} both as an lmo-based algorithm and as a trust-region algorithm.

>>>>

%------------------------------------------------------------------------------------
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
%\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{op}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{op}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

\usepackage{amsmath}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\diag}{diag}

@@

\subsection{Linear Minimization Oracle and Trust Region}
Let us look at the problem from the perspective of linear minimization oracle (lmo) and unconstrained stochastic conditional gradient descent (uSCG) (\cite{pethick2025training}). lmo is defined as:
\begin{equation}\label{eq:lmo}
    \mathrm{lmo}(\mS) \in \arg\min_{\mX \in \cS} \<\mS, \mX>,
\end{equation}
where $\cS$ is some set. We are interested in the case when $\cS$ is a ball in our $\norm{\cdot}$ norm:
\begin{equation}\label{eq:ball}
    \cS := \cB_\eta :=\{ \mX \mid \norm{\mX} \leq \eta \}.
\end{equation}

@@

It can be easily shown that the formula is equvalent to
\begin{equation}\label{eq:almost_trust_update}
    \mX^{k+1} = \mX^{k} - \gamma_k\eta \arg\max_{\mX \in \cB_1} \<\mS, \mX> = \mX^{k} - \gamma_k\eta \{\Delta \in \cB_1 \mid \<\mM^k, \Delta> = \norm{\mM^k}^\dagger\}.
\end{equation}

Let us set $\gamma_k = 1$. This transforms algorithm defined by \cref{eq:almost_trust_update} into Algorithm 1 from \cite{kovalev2025understanding}. Therefore, we can view \cref{eq:almost_trust_update} both as an lmo-based algorithm and as a trust-region algorithm.

\section{Different Norms $\norm{\cdot}$ imply different updates}
% As we have stated in Introduction, we are interested foremost in matrix norms, so we'll start right with them, skipping RMS-RMS norm and others.

Based on different norms $\norm{\cdot}$, we simplify the update defined by the aforementioned equation:

\begin{equation}\label{eq:our_update}
    \mX^{k+1} = \mX^{k} - \eta \{\Delta \in \cB_1 \mid \<\mM^k, \Delta> = \norm{\mM^k}^\dagger\}
\end{equation}

In all the work, we define $\mU$, $\mSigma$, $\mV^\top$ as components of the singular value decomposition of $\mM^k$: $\mM^k = \mU \mSigma \mV^\top$. We use common notations: $\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \sigma_2, \dots, \sigma_r)$, and  $\mV = [v_1, v_2, \dots, v_r]$.

{\bf $\normf{\mM^{k}}$ and NSGD}

Lemma. When $\norm{\cdot} = \normf{\cdot}$, \cref{eq:our_update} turns into:

\begin{equation}\label{eq:nsgd_update}
    \mX^{k+1} = \mX^{k} - \eta \frac{\mM^k}{\normf{\mM^k}}
\end{equation}

It is an interesting observation, because in other works (\cite{pethick2025training}), $\normf{\cdot}$ was used to recover SGD. The difference is in how one states the problem.

{\bf $\norms{\mM^{k}}$ and Muon}

Lemma. When $\norm{\cdot} = \norms{\cdot}$, \cref{eq:our_update} turns into:

\begin{equation}\label{eq:muon_update}
    \mX^{k+1} = \mX^{k} - \eta U V^\top
\end{equation}

{\bf $\normn{\mM^{k}}$ and Neon}

Lemma. When $\norm{\cdot} = \normn{\cdot}$, \cref{eq:our_update} turns into:

\begin{equation}\label{eq:neon_update}
    \mX^{k+1} = \mX^{k} - \eta u_1 v_1^\top
\end{equation}

We name the derived algorithm \emph{Neon}.

{\bf $\normfstar{\mM^{k}}^\dagger$ and F2-Muon}
We define $\normfstar{\cdot}$ as a convex combination of $\normn{\cdot}$ and $\normf{\cdot}$:

\begin{equation}\label{eq:fstar_norm}
    \normfstar{\mX} = \alpha \normn{\mX} + (1-\alpha)\normf{\mX},
\end{equation}

where $\alpha \in [0, 1]$ defines a specific norm of $F*$-family.

Lemma. When $\norm{\cdot} = \normfstar{\cdot}^\dagger$, \cref{eq:our_update} turns into:

\begin{equation}\label{eq:nsgd_muon_update}
    \mX^{k+1} = \mX^{k} - \eta (\alpha U V^\top + (1-\alpha)\frac{\mM^k}{\normf{\mM^k}})
\end{equation}

We name the derived algorithm \emph{F2-Muon}. It turns out that F2-Muon is a convex combination of Normalized SGD and Muon, which is curious. The implications are significant and discussed in the following sections.

{\bf $\normftwo{\mM^{k}}^\dagger$ and F2-Neon}
We define $\normftwo{\cdot}$ as a convex combination of $\norms{\cdot}$ and $\normf{\cdot}$:

\begin{equation}\label{eq:ftwo_norm}
    \normftwo{\mX} = \alpha \norms{\mX} + (1-\alpha)\normf{\mX},
\end{equation}

where $\alpha \in [0, 1]$ defines a specific norm of $F2$-family.

Lemma. When $\norm{\cdot} = \normftwo{\cdot}^\dagger$, \cref{eq:our_update} turns into:

\begin{equation}\label{eq:nsgd_neon_update}
    \mX^{k+1} = \mX^{k} - \eta (\alpha u_1 v_1^\top + (1-\alpha)\frac{\mM^k}{\normf{\mM^k}})
\end{equation}

We name the derived algorithm \emph{F2-Neon}. It turns out that F2-Neon is a convex combination of Normalized SGD and Neon, which is curious. The implications are significant and discussed in the following sections.

TO-DO: F*-neon? (it would be hard to do it)
